
\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}
\textit{Foremost, I would like to express my sincere gratitude to my principal advisor, Dr. Mirco Giacobbe, and my co-advisor, Dr. Leonardo Stella, for their continuous support of my research, as well as their patience, responsibility, motivation, and enthusiasm.}

\textit{Equally important, I would like to thank Dr. Pablo Samuel Castro, the author of the MICo paper, for his invaluable advice and guidance, which helped me to better understand the topic. I truly appreciate your prompt responses to my questions and your enthusiasm in explaining your work to me. I hope to collaborate with you on future projects.}

\textit{I would also like to thank my friends and colleagues who offered ideas and comments about my project. In particular, I want to highlight the observations of Fernando Zhapa, Anthony Ramos, Jose Seraquive, and Joseph Gonzalez, which were instrumental in shaping this project.}

\begin{CJK}{UTF8}{bsmi}
\textit{A special thank you goes to my dear friend Yu Wen (黃郁雯), who accompanied me every day in the library throughout the entire journey of this thesis. Thank you for feeding me and cheering me up with your weird situations (like "losing a laptop"). Without a doubt, you made this thesis a more enjoyable and pleasant experience, and I am deeply grateful for your company.}
\end{CJK}

\textit{Last but not least, I would like to express my heartfelt thanks to my family, who always motivated and supported me throughout this master's journey. I am particularly grateful to my parents, Elizabeth and Vicente, for their constant calls and for taking the time to talk to me during both challenging and happy moments. I also thank my sisters, Evelyn and Ericka, for their frequent video calls filled with stories that lifted my spirits and kept me moving forward. Without their unwavering support, I would not have successfully completed this stage of my life.}